---
title: "Antidepressant Medications Pre vs Post COVID19 - (S)ARIMA Modelling on individual data"
output:
  html_notebook: default
  html_document:
    df_print: paged
  pdf_document: default
---

```{r}
if(!require(tidyverse)) install.packages("tidyverse")
if(!require(zoo)) install.packages("zoo")
if(!require(tseries)) install.packages("tseries")
if(!require(fpp2)) install.packages("fpp2")


library(tidyverse)
library(tidyr)
library(dplyr)
library(ggplot2)
library(stringr)
library(lubridate)
library(zoo)
library(tseries)
library(fpp2)


```


```{r}

individual_prescriptions <- read_csv("Individual Monthly Prescriptions _ Analysis Ready.csv",
                          trim_ws = TRUE,
                          n_max = 106,
                          col_types = cols(MONTH_YEAR = col_date(format = "%b-%y")))

```

```{r}
str(individual_prescriptions)
```


Create new variables for MONTH & YEAR
```{r}
individual_prescriptions <- individual_prescriptions %>%
  mutate(
    YEAR = year(MONTH_YEAR),
    MONTH = month(MONTH_YEAR, label=T)
  )
```


```{r}
individual_prescriptions %>% 
  ggplot(aes(x=MONTH, y=n_antidepressants))+
  geom_boxplot()+
  labs(
    title = paste("Variation across Months in Individuals in Receipt of Antidepressant Prescriptions"),
    x = "Month",
    y = "Individuals (n)"
  )+
  theme_minimal()

```


```{r}
individual_prescriptions %>% 
  ggplot(aes(x=as.factor(YEAR), y=n_antidepressants))+
  geom_bar(stat="identity")+
  facet_wrap(~MONTH)+
  labs(
    title = paste("Yearly Trends for each Month"),
    x = "Month",
    y = "Individuals in receipt of antidepressants (n)"
  )+
  theme_minimal()

```



```{r}
individual_prescriptions %>% 
  ggplot(aes(x=MONTH_YEAR, y=n_antidepressants))+
  geom_line()+
  labs(
    title = paste("Trend in Prescriptions of antidepressant Medication"),
    y = "Individuals in receipt of antidepressants (n)"
  )+
  theme_minimal()
```



Split antidepressant data into PRE and POST COVID
```{r}
AD_precovid <- individual_prescriptions$n_antidepressants[1:98]      # PRECOVID Jan 2012 to Feb 2020
AD_postcovid <- individual_prescriptions$n_antidepressants[99:106]   # POSTCOVID Mar 2020 to Oct 2020

```


Create time series for precovid antidepressant data
```{r}

AD_precovid_ts <- ts(AD_precovid, 
                       start=c(2012,1),  #first in time series
                       freq=12)          # i.e. monthly data

```



Plot of monthly antidepressant prescriptions over the Precovid time frame
```{r}

autoplot(AD_precovid_ts) + ggtitle("Time Plot: Individuals in Receipt of Antidepressant Prescriptions per Month")+
  ylab("Number of Individuals in Receipt of Prescriptions")

```



Take first difference in data to remove upward trend and dip in 2017/18
```{r}
diff_AD_precovid <- diff(AD_precovid_ts)

```



Plot the differenced values (i.e. the month to month differences in prescription numbers)
```{r}
autoplot(diff_AD_precovid) + ggtitle("Time Plot: Monthly Change in Individuals in Receipt of antidepressant Prescriptions")+
  ylab("Number of Prescriptions")

```



Check for seasonality in stationary data set (i.e. the differenced data)
```{r}
ggseasonplot(diff_AD_precovid) + ggtitle("Seasonal Plot: Change in Monthly Prescriptions")+
  ylab("Number of Individuals")

```


Alternative view of seasonal variation
```{r}

ggsubseriesplot(diff_AD_precovid) + ggtitle("Seasonal Plot: Change in Monthly Prescriptions")+
  ylab("Number of Individuals")

```




## Model Building


1. Benchmark method: Seasonal naiive method
i.e. Each months value is equal to the next +/- random error
```{r}
fit <-snaive(diff_AD_precovid)  # Use differenced data (i.e. the un-trended data, "diff_AD_freq")

summary(fit)          

checkresiduals(fit)   
```




2. Exponential smoothing model
```{r}
fit_ets <- ets(AD_precovid_ts)     # ets() tries many ETS models and chooses the best performing
                                   # use the actual (not differenced) data
summary(fit_ets)

checkresiduals(fit_ets)

```



3. ARIMA model
```{r}

fit_arima <- auto.arima(AD_precovid_ts,       # Requires stationary data 
                        d=1,                  # d1 to take first difference of data to remove trend
                        D=1,                  # D1 to take the first seasonal difference to remove seasonal trends
                        stepwise = F,         # if true, the algorithm would try a smaller number to save time (costs accuracy)
                        approximation = F,    # if true, the AIC would be an approximation to save time (costs accuracy)
                        trace = T             # prints models as they run
                        )



```




Check performance of ARIMA on the precovid data
```{r}
summary(fit_arima)
checkresiduals(fit_arima)

```



Use the ARIMA model to forecast data into POSTCOVID timeframe
```{r}
forecast_arima <- forecast(fit_arima,
                           h=8) # forecast 8 months (March 2020 to Oct 2020)

autoplot(forecast_arima) # additional argument include = 12, would include only the last 12 months in precovid time frame

```


```{r}
summary(forecast_arima)

```


Extract the forecast values with the upper and lower 80 and 95% CI limits, and view the actual values for the POSTCOVID time frame
```{r}
forecast_arima_df <- data.frame(forecast_arima$mean)
lower_df <- as.data.frame(forecast_arima$lower,row.names = FALSE)
upper_df <- as.data.frame(forecast_arima$upper,row.names = FALSE)
AD_postcovid_df <- as.data.frame(AD_postcovid)


forecast_arima_df 
lower_df 
upper_df 
AD_postcovid_df 

```



Pull the values from forecast and actual values together in a data frame
```{r}
forecast_arima_df <- cbind(forecast_arima_df, lower_df, upper_df, AD_postcovid_df)

column_names <-  c("mean", "lower80", "lower95", "upper80", "upper95", "actual")

colnames(forecast_arima_df) <- column_names

MONTH <- c("Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")
forecast_arima_df <- forecast_arima_df %>%
  add_column(MONTH, .before = "mean")

forecast_arima_df$MONTH <- factor(forecast_arima_df$MONTH, levels = c("Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct"))

forecast_arima_df
```



Visualise the forcast prescription rates (with confidence intervals) versus the actual prescription rates in the POSTCOVID timeframe
```{r}
ggplot(data = forecast_arima_df,
       aes(x=forecast_arima_df$MONTH,y=mean, group=1))+
  geom_point()+
  geom_line()+
  geom_ribbon(aes(ymin=forecast_arima_df$lower95, ymax=forecast_arima_df$upper95), linetype=2, alpha=0.1)+
  geom_ribbon(aes(ymin=forecast_arima_df$lower80, ymax=forecast_arima_df$upper80), linetype=2, alpha=0.2)+
  geom_point(aes(x=forecast_arima_df$MONTH, y=forecast_arima_df$actual, group=1, col="red"))+
  geom_line(aes(x=forecast_arima_df$MONTH, y=forecast_arima_df$actual, group=1, col="red"))+
  labs(
    title = paste("Antidepressant medication prescriptions immediately after COVID-19 lockdown"),
    subtitle = paste("forecast vs. actual values"),
    caption = paste("black = forecast values\nred = actual values\ndark grey band = 80% confidence interval\nlight grey band = 95% confidence interval"),
    x = "Month",
    y = "Antidepressant prescriptions (n)"
  )+
  theme_minimal()+
  theme(legend.position = "none")
```



## MODEL ON DATA SET WITH 2017/18 REMOVED

Replace 2017 and 2018 data with "NA"
```{r}

AD_precovid_missing <- AD_precovid
AD_precovid_missing[61:84] <- NA      # 61=Jan 2017, 84=Dec2018
AD_precovid_missing

```



Take first difference in data to remove upward trend
```{r}

diff_AD_precovid_missing <- diff(AD_precovid_missing_ts)

```



1. Benchmark method: Seasonal naiive method
i.e. Each months value is equal to the next +/- random error
```{r}

fit_missing_AD <-snaive(diff_AD_precovid_missing)  # Use differenced data (i.e. the un-trended data, "diff_AD_freq")

summary(fit_missing_AD)          ## residual SD = 1521.5445  

checkresiduals(fit_missing_AD)   ## residuals over time look to be random and normally distributed- good
                      ## most error terms fall within 95% CI, one large at lag1


```




2. Exponential smoothing model
```{r}



fit_ets_missing_AD <- ets(AD_precovid_missing_ts)     # ets() tries many ETS models and chooses the best performing
                                           # use the actual (not differenced) data, as it can use trends
summary(fit_ets_missing_AD)

checkresiduals(fit_ets_missing_AD)

```




3. ARIMA model
```{r}

fit_arima_missing_AD <- auto.arima(AD_precovid_missing_ts,       # Requires stationary data 
                        d=1,                          # d1 to take first difference of data to remove trend
                        D=1,                          # D1 to take the first seasonal difference to remove seasonal trends
                        stepwise = F,                 # if true, the algorithm would try a smaller number to save time (costs accuracy)
                        approximation = F,            # if true, the AIC would be an approximation to save time (costs accuracy)
                        trace = T                     # prints models as they run
                        )



```



Check performance of ARIMA on the precovid data
```{r}
summary(fit_arima_missing_AD)
checkresiduals(fit_arima_missing_AD)

```



Use the ARIMA model (with missing values) to forecast data into POSTCOVID timeframe
```{r}
forecast_arima_missing_AD <- forecast(fit_arima_missing_AD,
                           h=8) # forecast 8 months (March 2020 to Oct 2020)

autoplot(forecast_arima_missing_AD) # additional argument include = 12, would include only the last 12 months in precovid time frame

```


```{r}
summary(forecast_arima_missing_AD)

```




Extract the forecast values with the upper and lower 80 and 95% CI limits, and view the actual values for the POSTCOVID time frame
```{r}
forecast_arima_df_missing_AD <- data.frame(forecast_arima_missing_AD$mean)
lower_df_missing_AD <- as.data.frame(forecast_arima_missing_AD$lower,row.names = FALSE)
upper_df_missing_AD <- as.data.frame(forecast_arima_missing_AD$upper,row.names = FALSE)
postcovid_df_AD <- as.data.frame(AD_postcovid)


forecast_arima_df_missing_AD 
lower_df_missing_AD
upper_df_missing_AD
postcovid_df_AD 

```




Pull the values from forecast and actual values together in a data frame
```{r}
forecast_arima_df_missing_AD <- cbind(forecast_arima_df_missing_AD, lower_df_missing_AD, upper_df_missing_AD, postcovid_df_AD)

column_names <-  c("mean", "lower80", "lower95", "upper80", "upper95", "actual")

colnames(forecast_arima_df_missing_AD) <- column_names

MONTH <- c("Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct")
forecast_arima_df_missing_AD <- forecast_arima_df_missing_AD %>%
  add_column(MONTH, .before = "mean")

forecast_arima_df_missing_AD$MONTH <- factor(forecast_arima_df_missing_AD$MONTH, levels = c("Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct"))

forecast_arima_df_missing_AD
```




Visualise the forcast prescription rates (with confidence intervals) versus the actual prescription rates in the POSTCOVID timeframe
```{r}
ggplot(data = forecast_arima_df_missing_AD,
       aes(x=forecast_arima_df_missing_AD$MONTH,y=mean, group=1))+
  geom_point()+
  geom_line()+
  geom_ribbon(aes(ymin=forecast_arima_df_missing_AD$lower95, ymax=forecast_arima_df_missing_AD$upper95), linetype=2, alpha=0.1)+
  geom_ribbon(aes(ymin=forecast_arima_df_missing_AD$lower80, ymax=forecast_arima_df_missing_AD$upper80), linetype=2, alpha=0.2)+
  geom_point(aes(x=forecast_arima_df_missing_AD$MONTH, y=forecast_arima_df_missing_AD$actual, group=1, col="red"))+
  geom_line(aes(x=forecast_arima_df_missing_AD$MONTH, y=forecast_arima_df_missing_AD$actual, group=1, col="red"))+
  labs(
    title = paste("Individuals in Receipt of Antidepressant Medication immediately after COVID-19 lockdown"),
    subtitle = paste("forecast vs. actual values"),
    caption = paste("black = forecast values\nred = actual values\ndark grey band = 80% confidence interval\nlight grey band = 95% confidence interval"),
    x = "Month",
    y = "Individuals (n)"
  )+
  theme_minimal()+
  theme(legend.position = "none")
```

